---
title: shiny app as R package
author: Ildi Czeller
date: '2017-08-09'
slug: shiny-app-as-r-package
categories: ["R"]
tags: ["shiny", "package"]
---

### Background

I love and use shiny regularly and I also try to modularize my work more and more. Naturally I started to wonder whether I could have a whole shiny app as an R package and then possibly a project organizing multiple apps together. The answer is yes as I suspected.

Several others already wrote about the topic, including [Dean Attali](https://www.r-bloggers.com/supplementing-your-r-package-with-a-shiny-app-2/) and [William Landau](http://will-landau.com/2016/11/01/appPackage/). However, I gained a deeper understanding by trying things out myself.

You have two main options to run a shiny app:

 - either launch shiny in a directory with a `server.R` and `ui.R` file,
 - or launch a shiny object with ui and server parts defined appropriately.

Until now I always used the first method but to put an app in a package you need the second. This is probably why it took some time for me to understand how a package containing a shiny app can work.

### Implementation

Taking this functional method there is actually no difficulty in putting a shiny app in a package. You define a function for both your ui and server parts and have a function for launching the app which is the only function you need to export from your package.

```{r, eval = FALSE}
shinypackage_ui <- function() {
    fluidPage(...)
}

shinypackage_server <- function(input, output, session) {
    ...
}
```

```{r, eval = FALSE}
#' @param ... passed to runApp, e.g. port, launch.browser
#' @export
runShinyPackageApp <- function(...) {
    app <- shinyApp(
        ui = shinypackage_ui(),
        server = shinypackage_server
    )
    runApp(app, ...)
}
```

If you have data loading during runtime or any dependencies outside your package it will work as usual.

### (Not) using `global.R`

One significant difference between the two methods to run a shiny app is the use of the `global.R` file. (The `onStart` param is related but different.) I generally use `global.R` for the following things:

* source libraries,
* source functions,
* define project-wide constants or load data used in `ui`,
* define values which may change between launches of the app.

The first three points are in fact easier and cleaner in a package.

* You can import packages package-wide or for specific functions,
* all functions within package are automatically available to other functions,
* you can use internal data of any kind in your package.

For the last one instead of declaring global variables (not a good practice anyways) you may pass paramaters to the  `runShinyPackageApp` function which can pass them on to the server and ui, for example:

```{r, eval = FALSE}
runShinyPackageApp <- function(launch_param = NULL, ...) {
    app <- shinyApp(
        ui = shinypackage_ui(launch_param),
        server = shinypackage_server
    )
    runApp(app, ...)
}
```

### Conclusion

I recommend putting shiny apps in R packages but even if I don't I will use the functional form more as it seems clearer to me now.

You can see the the whole code on [github](https://github.com/czeildi/shinypackage) and try for yourselves with:

```{r, eval = FALSE}
devtools::install_github(
    "czeildi/shinypackage", ref = "v0.1.0"
)
```
