---
title: merge on factor column
author: Ildi Czeller
date: '2017-11-30'
slug: merge-on-factor-column
categories: ["R"]
tags: ["forcats"]
---

```{r setup, echo =FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


```{r}
library("forcats")
library("data.table")
library("magrittr")
```

I recently started working with factors in R and found the forcats package amazing. However, I did not found much documentation on merging two data frames on a factor column so I made small examples and observed the behavior.

## same levels, different order

In abc order: natural if converting from already ordered character.

```{r}
dt_abc <- data.table(size = as_factor(c("big", "medium", "tiny")))
levels(dt_abc$size)
```

We can define the natural order: 
```{r}
dt_lg <- data.table(size = fct_reorder(as_factor(c("big", "medium", "tiny")), c(3, 2, 1)))
levels(dt_lg$size)
```

```{r}
merge(dt_abc, dt_lg, by = "size") %T>%
{print(levels(.$size))}
```
```{r}
merge(dt_lg, dt_abc, by = "size") %T>%
    {print(levels(.$size))}
```

From this it seems that with merge the left table has precedence.

## different levels

```{r}
dt_abc <- data.table(size = as_factor(c("big", "medium", "small", "tiny")))
dt_lg <- data.table(size = fct_reorder(as_factor(c("big", "medium", "tiny")), c(3, 2, 1)))
```

```{r}
merge(dt_abc, dt_lg, by = "size", all = TRUE) %T>%
    {print(levels(.$size))}
```

```{r}
merge(dt_lg, dt_abc, by = "size", all = TRUE) %T>%
    {print(levels(.$size))}
```
In this case the levels of the factors in the two table are concatenated: first left then right.

## merge factor with character

```{r}
dt_chr <- data.table(size = c("big", "medium", "tiny"))
dt_fct <- data.table(size = fct_reorder(as_factor(c("big", "medium", "tiny")), c(3, 2, 1)))
```

```{r}
merge(dt_fct, dt_chr, by = "size") %T>%
    {print(levels(.$size))}
```
```{r}
merge(dt_chr, dt_fct, by = "size") %T>%
    {print(levels(.$size))}
```

From this it seems that in this case the right side of the join has precedence.
