---
title: shiny app as R package v2
author: Ildi Czeller
date: '2017-08-16'
slug: shiny-app-as-r-package-v2
categories: ["R"]
tags: ["shiny", "package"]
---

---
output: 
  html_document:
    css: styles.css
---

In my previous [post](https://www.ildiczeller.com/2017/08/09/shiny-app-as-r-package/) I explored the possibility of wrapping a shiny app as an R package. I falsely stated that you cannot have a directory-based shiny app with `global.R`. Based on [this](https://stackoverflow.com/questions/37830819/developing-shiny-app-as-a-package-and-deploying-it-to-shiny-server) question on Stack Overflow I proved the contrary. It is actually possible but with serious limitations in my opinion so I still strongly advise the function based approach.

You can put basically anything in the `inst` folder of your package and use those as is. I.e. you can put your regular `server.R`, `ui.R` and `global.R` files to `inst` and have an exported launcher function in `R`:

```{r eval = FALSE}
#' @export
runShinyPackageApp <- function(...) {
    shiny::runApp(
        appDir = system.file(".", package = "shinypackageapp"),
        ...
    )
}
```


## Obstacles I ran into

* You cannot simply use functions from imported packages unless you explicitly call `library(dependecy_package)` in `server.R` or `global.R`.
* If you create utility functions you cannot used them unless they are exported and you call `library(shinypackageapp)` before `runShinyPackageApp()`.
* You cannot access internal package data in the `server.R` or `ui.R` files unless you have an exported function to access your package data.
* You cannot access the file system regularly, if you check the contents of your default folder in `server.R` you will see something similar:

```
DESCRIPTION
global.R
help
html
INDEX
LICENSE
Meta
NAMESPACE
R
server.R
ui.R
```

I suspect this is some kind of temp location of the installed package, certainly not the root folder of my project, neither the root folder of the package as I see it during package development.

All the aboves are due to the fact that these files are not compiled in any way but partly pretend to exist in the folder where you installed the package.

## Conclusion

Besides distribution you loose most of the positive sides of wrapping your app in a package this way.

You can see the the whole code on [github](https://github.com/czeildi/shinypackageapp) and try for yourselves with:

```{r, eval = FALSE}
devtools::install_github(
    "czeildi/shinypackageapp", ref = "v0.4.0"
)
```
