---
title: Investigating difftime behavior
author: Ildi Czeller
date: '2018-04-02'
slug: investigating-difftime-behavior
categories: ["R"]
tags: ["base R", "S3", "lookup"]
---

*Author: Ildi Czeller, @czeildi on Twitter and Github*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
knitr::knit_hooks$set(
  source = highlight::hl_hook_source,
  document = highlight::hl_hook_css
)
```

```{r echo=FALSE}
suppressPackageStartupMessages(library("data.table"))
```


It caused me a great deal of headache to work around `as.difftime` to get what I wanted but also I learned much more than I expected along the way so I wanted to share the journey with you.

## Initial experience

I had a data frame with two columns containing timestamps and I wanted to calculate the difference between them.

```{r echo=FALSE}
dt <- data.table(
    send_time = as.POSIXct(
        c("2018-04-01 00:01:00", "2018-04-02 00:01:00")
    ),
    open_time = as.POSIXct(
        c("2018-04-01 00:07:00", "2018-04-02 02:01:00")
    )
)
dt
```

In the first row, there are 6 minutes between send and open time and in the second row there are 2 hours. The essence of the unexpected behavior I experienced:

```{r}
dt[, open_time - send_time][2]
```

```{r}
dt[2, open_time - send_time]
```

Basically the result for one row depends on other rows in the data frame.

### Finding a workaround

My first idea was to explicitly provide the `units` argument. However, it did not help.

```{r}
dt[, as.difftime(open_time - send_time, units = "hours")][2]
```

I found that if I create a difftime object with `difftime` and not with subtraction it works as I expect.

```{r}
dt[, difftime(open_time, send_time, units = "hours")][2]
```

## Digging deeper

### How does timestamp subtraction actually work? 

```{r}
base_time <- lubridate::ymd_hms("2018-04-01 00:00:00")
later_times <- base_time + c(
    lubridate::period(1, "month"),
    lubridate::period(1, "day"),
    lubridate::period(1, "minute"),
    lubridate::period(1, "second")
)
```

```{r}
later_times[1:1] - base_time
```
```{r}
later_times[1:2] - base_time
```
```{r}
later_times[1:3] - base_time
```
```{r}
later_times[1:4] - base_time
```

Based on this I am quite confident to say that firstly the smallest unit will be used for all values and secondly subtraction won't result in units greater than a day.

### Why does `units` not help? 

It still puzzled me why the explicit `units` argument did not help. Time to look at the source code at last!

```{r, eval=FALSE}
as.difftime
```

```{r, eval=FALSE}
function (tim, format = "%X", units = "auto")
{
    if (inherits(tim, "difftime"))
        return(tim)
    if (is.character(tim)) {
        difftime(strptime(tim, format = format), strptime("0:0:0",
            format = "%X"), units = units)
    }
    else {
        # ...
    }
}
```


Only by skimming the code we can see that it behaves differently depending on the type of its argument. What did we pass? 

```{r}
class(dt[2, open_time - send_time])
```

In the first two rows we get our answer: if the argument to `as.difftime` already a `difftime` object, then the `units` argument will have no effect.

### `units="auto"`

My next question was where is the code to determine the used unit? As `as.difftime` makes no transformation it must happen when we subtract two vectors.

Now it is useful to know that `-` is an S3 generic which means it selects what S3 method to call based on the type (S3 class) of its argument(s). In base R subtraction is defined specifically for timestamp objects (of class `POSIXt`). Note that you can call S3 methods directly with `[generic_name].[class_name]`.

```{r, eval=FALSE}
`-.POSIXt`
```

```{r, eval=FALSE}
function (e1, e2)
{
    # ...
    if (!inherits(e1, "POSIXt"))
        stop("can only subtract from \"POSIXt\" objects")
    if (nargs() == 1)
        stop("unary '-' is not defined for \"POSIXt\" objects")
    if (inherits(e2, "POSIXt"))
        return(difftime(e1, e2))
    # if ...
    # ...
}
```


We subtract two values of class `POSIXt` so the relevant part for us is `return(difftime(e1, e2))`. Let's look at `difftime` now.

```{r, eval=FALSE}
difftime
```

```{r, eval=FALSE}
function (time1, time2, tz, units = c("auto", "secs", "mins",
    "hours", "days", "weeks"))
{
    # ...
    z <- unclass(time1) - unclass(time2)
    attr(z, "tzone") <- NULL
    units <- match.arg(units)
    if (units == "auto")
        units <- if (all(is.na(z)))
            "secs"
        else {
            zz <- min(abs(z), na.rm = TRUE)
            if (!is.finite(zz) || zz < 60)
                "secs"
            else if (zz < 3600)
                "mins"
            else if (zz < 86400)
                "hours"
            else "days"
        }
    # ...
}
```

Without understanding every detail we can see that if `units` are not specified as in our case it will depend on the smallest absolute value. What makes these computations work is that `unclass()` on a `POSIXct` type object gives the number of seconds by default since `1970-01-01`.

Finally I took the time to read the longish help file of `difftime` which contained most of this information. Still I would go this way again as I believe I learned more.

### Conclusions

Look at the source code earlier and always be explicit with conversions if possible.
